A simulation model of DNA methylation data

### 2014-10-17 Rmarkdown/methsim.Rmd
Reviewed methods for simulating BS-seq reads

@@

>>>>

Simulated bisulfite-sequencing data may be used in the comparison of strategies for the alignment and processing of bisulfite-converted reads, or in the comparison of downstream analysis methods. Simulation methods for bisulfite-sequencing data may choose to simulate the individual sequencing reads, unaligned or aligned, or some summary measure of methylation, such as $\beta$-values. The former is required for the comparison of different alignment strategies whereas the latter is typically used by developers of downstream analysis methods, such as differential methylation callers, since this is often all that is required to run these methods. Moreover, simulated reads must be aligned and processed before they can be used in comparisons of downstream analysis methods, which is often not of direct relevance in comparisons of these tools.

## Methods for simulating bisulfite-sequencing reads

All of the available methods for simulating bisulfite-sequencing reads are designed for the comparison of alignment strategies. These are not suitable for comparisons of downstream analysis methods.

`Sherman` ([http://www.bioinformatics.babraham.ac.uk/projects/sherman/](http://www.bioinformatics.babraham.ac.uk/projects/sherman/)) is software to simulate bisulfite-sequencing reads, including various 'contaminants', such as SNPs, basecall errors and sequence artefacts as `FASTQ` files. The simulated reads are designed for comparing the performance of different alignment strategies. It has many parameters, of which the ones relevant to our discussion of simulating realistic DNA methylation data are `-CG` and `-CH`, the bisulfite conversion rates for CG and CH methylation loci, respectively. These are set by the user with values between $0$ and $100$. Reads are simulated by sampling from the user-specified reference genome. When a read contains a CG (resp. CH) locus, it is randomly assigned as being converted (unmethylated) with probability `-CG` / 100 (resp. `-CH` / 100).

While suitable for comparing alignment strategies, `Sherman` produces data that is not suitable for use in comparing downstream analysis methods. All CG (resp. CH) loci have an average $\beta$-value of `-CG` (resp. `-CG`) regardless of the genomic context, which we know to be inconsistent with real data. Furthermore, the methylation state of each methylation locus is independent, which is clearly inconsistent with the strong co-methylation observed in real data.

`FastqToBS` ([http://users.dimi.uniud.it/~nicola.prezza/projects.html](http://users.dimi.uniud.it/~nicola.prezza/projects.html)) uses the same strategy as `Sherman`.

`DNemulator` ([http://www.cbrc.jp/dnemulator/README.html](http://www.cbrc.jp/dnemulator/README.html)) uses a slightly more sophisticated simulation strategy to simulate `FASTQ` files for use in comparing alignment strategies for bisulfite-sequencing data. `DNemulator` does this with three separate routines, `fasta-methly-sim`, `fasta-polymorph` and `fasta-bisulf-sim`:

1. `fasta-methyl-sim` converts cytosines in the reference genome (`FASTA` file) to a character indicating the methylation level of that locus: `C` represents $0\%$ methylated, `c` represents $10\%$ methylated, `d` represents $20\%$ methylated, `v` represents $50\%$ methylated and `t` represents $100\%$ methylated. Each of these conversions has a different probability in the CG and CH contexts.
2. `fasta-polymorph` simulates a polymorphic, diploid genome based on the modified reference sequence created by `fasta-methyl-sim`.
3. `fasta-bisulf-sim` simulates reads by sampling from the simulated genome created by `fasta-polymorph`. Read are simulated with bisulfite-conversion error and sequencing error.

The reads simulated by `DNemulator` will result in $\beta$-values that have more context-dependence than those resulting from reads generated by `Sherman`, `FastqToBS` or `BSsim`. However, methylation events are still generated independently of one another, which means there is no co-methylation in the simulated data. Therefore, reads simulated by `DNemulator`, while suitable for comparing alignment strategies, are not suitable for comparing downstream analysis methods.

`BSsim` ([http://122.228.158.106/BSSim/](http://122.228.158.106/BSSim/) and used in \cite{Xie:2014ez}), uses a similar strategy to `DNemulator`.

## Methods for $\beta$-values

Methods to describe:

- [ ] \cite{Feng:2014iq}
- [ ] \cite{Lacey:2013iy}
- [ ] \cite{Sofer:2013bk}
- [ ] \cite{Xu:2013eg}
- [ ] \cite{Chen:2014jb}
- [ ] \cite{Park:2014ho}
- [ ] \cite{Dolzhenko:2014bo}
- [ ] \cite{Xie:2014ez}
- [ ] Others?

@@

`methsim` is designed for simulating data for downstream analyses, and uses a hybrid strategy that is capable of creating raw, unaligned reads or summary measures, such as $\beta$-values, from the same simulation run.
